---
title: "Coursera - Practical Machine learning assignment"
author: "Yashwanth"
date: "January 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This analysis is done as a part of Project Assignment for Practical Machine Learning course of the John Hopkins Data Science Specialization at Coursera. 
The project uses data from the Weight Lifting Exercises (WLE) Dataset (see http://groupware.les.inf.puc-rio.br/har and also the References section below.).
According to the WLE website, six participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions, identified as classes A, B, C, D and E. Class A corresponds to a correct execution of the exercise, and the remaining classes identify common mistakes in this weight lifting exercise. Several sensors were used to collect data about the quality of the exercise execution. 

The goal of this project is to develop a machine learning algorithm that takes a set of sensor readings and accurately predicts the corresponding class (A to E).

The following analysis uses a random forest prediction algorithm to accomplish this task, after some data cleaning. The results of the analysis confirm that the model provided by this algorithm achieves a high prediction accuracy (as indicated by several prediction quality indicators).


###Loading required libraries

```{r loading_library}

library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)

```


###Loading the Train and Test Data

```{r loading_file}
set.seed(12345)
train_data_Url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_data_Url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training <- read.csv(url(train_data_Url), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(test_data_Url), na.strings=c("NA","#DIV/0!",""))

```




### Splitting train dataset further into Train and Test datasets

```{r splitting}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
Train_subtrain <- training[inTrain, ]; Train_subtest <- training[-inTrain, ]
dim(Train_subtrain)
dim(Train_subtest)

```

### Cleaning the data

```{r cleaning}
myDataNZV <- nearZeroVar(Train_subtrain, saveMetrics=TRUE)
myNZVvars <- names(Train_subtrain) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
                                      "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
                                      "max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
                                      "var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
                                      "stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
                                      "kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
                                      "max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
                                      "kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
                                      "skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
                                      "amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
                                      "skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
                                      "max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
                                      "amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
                                      "avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
                                      "stddev_yaw_forearm", "var_yaw_forearm")
Train_subtrain <- Train_subtrain[!myNZVvars]
dim(Train_subtrain)


```


###Updating NAs in both train and test datasets

```{r update_na}
Train_subtrain <- Train_subtrain[c(-1)]
trainingV3 <- Train_subtrain #creating another subset to iterate in loop
for(i in 1:length(Train_subtrain)) { #for every column in the training dataset
  if( sum( is.na( Train_subtrain[, i] ) ) /nrow(Train_subtrain) >= .6 ) { #if n?? NAs > 60% of total observations
    for(j in 1:length(trainingV3)) {
      if( length( grep(names(Train_subtrain[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
        trainingV3 <- trainingV3[ , -j] #Remove that column
      }
    }
  }
}
dim(trainingV3)
Train_subtrain <- trainingV3
rm(trainingV3)
clean1 <- colnames(Train_subtrain)
clean2 <- colnames(Train_subtrain[, -58])

Train_subtest <- Train_subtest[clean1]

testing <- testing[clean2]
dim(Train_subtest)
dim(testing)

```


###Making train and test datasets consistent with formats


```{r consistency}
for (i in 1:length(testing) ) {
  for(j in 1:length(Train_subtrain)) {
    if( length( grep(names(Train_subtrain[i]), names(testing)[j]) ) ==1)  {
      class(testing[j]) <- class(Train_subtrain[i])
    }
  }
}
testing <- rbind(Train_subtrain[2, -58] , testing)
testing <- testing[-1,]

```




###Developing Decision tree model
```{r decision_tree}
modFitA1 <- rpart(classe ~ ., data=Train_subtrain, method="class")
fancyRpartPlot(modFitA1)
```


###Predicting using test dataset and performing validation checks
```{r dt_prediction}
predictionsA1 <- predict(modFitA1, Train_subtest, type = "class")


confusionMatrix(predictionsA1, Train_subtest$classe)
```

###Developing Random forest
```{r random_forest}
modFitB1 <- randomForest(classe ~. , data=Train_subtrain)
predictionsB1 <- predict(modFitB1, Train_subtest, type = "class")
confusionMatrix(predictionsB1, Train_subtest$classe)
predictionsB2 <- predict(modFitB1, testing, type = "class")

```


###Writing output files to the desired location in local computer
```{r Results_export}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(predictionsB2)
```
